### 2018-03-15 paper/6-engineering.tex

\clearpage
\section{Engineering aspects}\label{sec-engineering}

@@

Our model assumes a world where the store is fully-defined, every \hs{k} is associated with a \hs{v}, and every compute successfully completes returning a valid value. In the real world, build systems frequently deal with errors - e.g. ``file not found'', or ``compilation failed''. We can model such failure conditions by instantiating \hs{v} to either \hs{Maybe v} (for missing values) or \hs{Either e v} (for exceptions of type \hs{e}). That can model the the values inside the store and the compute function, but because \hs{v} is polymorphic for builds, it does not let the build system apply special behaviour for errors - e.g. early aborting. We consider such details an engineering concern, which has no real impact on the power or high-level implementation of the build system.

@@

\todo{SLPJ}{Can we replace ``node'' with ``task''? Everywhere.}
\begin{enumerate}
\item The \hs{topological} function can instead retain the dependency graph, and whenever all the dependencies of a node are complete, the node itself can be started.
\item The \hs{reordering} function can be made parallel in a few ways, but the most direct is to have $n$ threads reading entries from the list of keys. As before, if a key requires something not yet built, it is added to the end -- the difference is that sometimes things will be moved back not because they are out of order but because of races with earlier nodes that had not yet finished. As a consequence, over successive runs, potentially racey dependencies will be separated, giving better parallelism over time.
\item The \hs{recursive} function can be made parallel by working more directly on the \hs{Depeneds} structure, which gives multiple dependency keys at a time (all those arising from \hs{Applicative} dependencies). All such dependencies can be executed simultaneously, with a slightly smarter check for whether something is already executing that also blocks until an existing execution has finished.
\end{enumerate}

@@

In our model we define \hs{Task} as a function - when give the same inputs it will always produce the same outputs. Alas, the real-world is not as obliging. Some examples of impure \hs{Task} functions include:

\begin{description}
\item[Untracked state] Some actions depend on untracked state - for example your C compilation may explicitly list the \texttt{source.c} file as a dependency, but it may not record that the version of \texttt{gcc} is also a dependency.
\item[Volatility] Some actions are defined to change in every execution, for example \Excel has a function \textsf{RANDBETWEEN} which produces a random number - on each recalculation this function is defined to change. Similarly, build systems like \Make and \Shake provide \textsf{phony} rules which are also volatile.
\item[Non-determinism] Some actions are non-determinstic, producing one result from a possble set. As an example, GHC when compiled using parallelism can change the order in which unique variables are obtained from the supply, producing semantically identical results but with different hidden variables.
\end{description}

@@

\subsection{Cloud aspects}\label{sec-cloud-aspects}

@@

\begin{description}
\item[Materialisation] Sometimes input files will go through many transformations before producing the end result, e.g. an installer package. These intermediate steps may be large, so some cloud builds can execute the graph \textit{without} downloading or \textit{materialising} the intermediate nodes - only the final one. Some build systems can go even further, integrating with the file system to only materialise the file when the user accesses it \cite{google_internal}.
\item[Eviction] The store of traces as shown grows indefinitely. In many cases resource constraints mean that eventually items must be evicted from the store that are predicted not to be used again. One option is to evict the contents and any trace that mentions the now-defunct \hs{Hash}. However, if the build system can defer materialisation, it may be possible to only evict the contents, allowing builds to pass-through hashes of values where the underlying value is not known. If so, the build must be able to recreate the value if required, potentially dealing with a different result in a future run.
\item[Frankenbuilds] A build is considered a franken-build \cite{cloudbuild} if a value is calculated locally, but something that depends on that key is pulled from the cache, and the value calculated locally does not match what was previously calculated by the cloud. Our implementations avoid this issue by storing complete traces, but if a cloud build system was to only reference input nodes it could be a problem.
\item[Communication] With a cloud build communication with a central server can be a bottleneck, so it is important to send only the minimum amount of information, optimising with respect to build-system specific invariants - see \S\ref{sec-smart-traces} for some possible optimisations.
\item[Offloading] Once the cloud is storing build products and traces, it is not infeasible for the cloud to also contain dedicated workers that can help with jobs run locally - offloading some of the computation and potentially running vastly more commands in parallel.
\end{description}

@@

Some build systems, for example \Excel and \Ninja are capable of recomputing a node if either its dependencies change, \textit{or} the rule itself changes. For example:

\begin{minted}[xleftmargin=10pt]{bash}
A1 = 20
A2 = 10
B1 = A1+A2
\end{minted}

In \Excel the user can alter the value produced by \textsf{B1} by either editing the inputs of \textsf{A1} or \textsf{A2}, \textit{or} editing the formula in \textsf{B1} - e.g. to \textsf{A1-A2}. This pattern can be captured by describing the rule producing \textsf{B1} as also depending on the value \textsf{B1.formula}. In a \hs{Monad} build system the implementation can be given very directly as:

\begin{minted}{haskell}
compute ask "B1" = do
    formula <- ask "B1-formula"
    evalFormula ask formula
\end{minted}

Namely, first look up the formula, then interpret it. In an \hs{Applicative} system it is not possible to change dependencies based on the formula, as would be required, so instead the formula can be captured as a dependency (but its value not used) and \textit{also} baked directly into the \hs{compute} function.

@@

Our build system abstraction assumes a \hs{k -> v} store, along with a build system that works directly on \hs{k} and \hs{v} values. However, certain build systems provide greater flexibility, e.g. \Shake permits polymorphic keys and values, allowing types that are only stored in the Shake information, and never persisted to the Store.

@@

Another more compelling example is build rules that produce multiple keys -- for example \textsf{ghc Foo.hs} produces both \textsf{Foo.hi} and \textsf{Foo.o}. That can be represented by having a key whose value is a pair of file names, and whose result is a pair of file contents. From that, the rule for \textsf{Foo.hi} can be the first-component of the result of the pair. Again, such an operation can be encoded without polymorphic keys provided the pair of files (or a dummy file representing the pair) is marked as changed if either of the contained files change. Once again, polymorphic dependencies provide convenience rather than power.

>>>>

\section{Engineering aspects}\label{sec-engineering}

@@

Our model assumes a world where the store is fully-defined, every \hs{k} is associated with a \hs{v}, and every compute successfully completes returning a valid value. In the real world, build systems frequently deal with errors - e.g. ``file not found'', or ``compilation failed''. We can model such failure conditions by instantiating \hs{v} to either \hs{Maybe v} (for missing values) or \hs{Either e v} (for exceptions of type \hs{e}). That can model the the values inside the store and the \hs{compute} function, but because \hs{v} is polymorphic for builds, it does not let the build system apply special behaviour for errors - e.g. early aborting.

@@

\begin{enumerate}
\item The \hs{topological} function can build the full dependency graph, and whenever all dependencies of a task are complete, the task itself can be started.
\item The \hs{reordering} function can be made parallel in a few ways, but the most direct is to have $n$ threads reading entries from the list of keys. As before, if a key requires something not yet built, it is added to the end -- the difference is that sometimes things will be moved to the back of the queue not because they are out of order but because of races with earlier nodes that had not yet finished. As a consequence, over successive runs, potentially racey dependencies will be separated, giving better parallelism over time.
\item The \hs{recursive} function can be made parallel by starting \hs{Applicative} dependencies of a \hs{Task} in parallel, using the strategy described by \citet{haxl}.
\end{enumerate}

@@

In our model we define \hs{Task} as a function - when give the same inputs it will always produce the same outputs. Alas, the real-world is not so obliging. Some examples of impure \hs{Task} functions include:

\begin{description}
\item[Untracked state] Some actions depend on untracked state - for example your C compilation may explicitly list the \texttt{source.c} file as a dependency, but it may not record that the version of \texttt{gcc} is also a dependency.
\item[Volatility] Some actions are defined to change in every execution, for example \Excel has a function \textsf{RANDBETWEEN} which produces a random number - on each recalculation this function is defined to change. Similarly, build systems like \Make and \Shake provide \textsf{phony} rules which are also volatile.
\item[Non-determinism] Some actions are non-determinstic, producing one result from a possble set. As an example, GHC when compiled using parallelism can change the order in which unique variables are obtained from the supply, producing different but semantically identical results.
\end{description}

@@

\subsection{Cloud implementations}\label{sec-cloud-aspects}

@@

\begin{description}
\item[Materialisation] Sometimes input files will involve many intermediate tasks before producing the end result, e.g. an installer package. These intermediate steps may be large, so some cloud builds can execute the graph \textit{without} downloading or \textit{materialising} the results of intermediate tasks - only the final result. Some build systems can go even further, integrating with the file system to only materialise the file when the user accesses it \cite{google_internal}.
\item[Eviction] The store of traces as shown grows indefinitely, but often resource constraints require evicting old items from the store. One option is to evict the contents and any trace that mentions the now-defunct \hs{Hash}. However, if the build system can defer materialisation, it may be possible to only evict the contents, allowing builds to pass-through hashes of values where the underlying value is not known. If so, the build must be able to recreate the value if required, potentially dealing with a different result in a future run.
\item[Frankenbuilds] A build is considered a franken-build \cite{cloudbuild} if a value is calculated locally, but something that depends on that key is pulled from the cache, and the value calculated locally does not match what was previously calculated by the cloud. Our implementations avoid this issue by storing complete traces, but if a cloud build system was to only reference input nodes this situation can arise.
\item[Communication] When traces or contents are stored on a central server communication can become a bottleneck, so it is important to send only the minimum amount of information, optimising with respect to build-system specific invariants - see \S\ref{sec-smart-traces} for some possible optimisations.
\item[Offloading] Once the cloud is storing build products and traces, it is possible for the cloud to also contain dedicated workers that can execute tasks remotely - offloading some of the computation and potentially running vastly more commands in parallel.
\end{description}

@@

Some build systems, for example \Excel and \Ninja, are capable of recomputing a task if either its dependencies change, \textit{or} the rule itself changes. For example:

\begin{minted}[xleftmargin=10pt]{text}
A1 = 20            B1 = A1+A2
A2 = 10
\end{minted}

In \Excel the user can alter the value produced by \textsf{B1} by either editing the inputs of \textsf{A1} or \textsf{A2}, \textit{or} editing the formula in \textsf{B1} - e.g. to \textsf{A1-A2}. This pattern can be captured by describing the rule producing \textsf{B1} as also depending on the value \textsf{B1-formula}. In a \hs{Monad} build system the implementation can be given very directly as:

\begin{minted}{haskell}
task fetch "B1" = do
    formula <- fetch "B1-formula"
    evalFormula fetch formula
\end{minted}

Namely, first look up the formula, then interpret it. In an \hs{Applicative} system it is not possible to change dependencies based on the formula, as would be required, so instead the formula can be captured as a dependency (but its value not used) and \textit{also} baked directly into the \hs{task} function.

@@

Our build system abstraction assumes a \hs{k}/\hs{v} store, along with a build system that works directly on \hs{k} and \hs{v} values. However, certain build systems provide greater flexibility, e.g. \Shake permits polymorphic keys and values, allowing types that are only stored in the Shake information, and never persisted to the Store.

@@

A more compelling example is build rules that produce multiple keys -- for example \textsf{ghc Foo.hs} produces both \textsf{Foo.hi} and \textsf{Foo.o}. That can be represented by having a key whose value is a pair of file names, and whose result is a pair of file contents. From that, the rule for \textsf{Foo.hi} can be the first-component of the result of the pair. Again, such an operation can be encoded without polymorphic keys provided the pair of files (or a dummy file representing the pair) is marked as changed if either of the contained files change. Once again, polymorphic dependencies provide convenience rather than power.

### 2018-03-15 paper/8-conclusions.tex

\clearpage
\section{Conclusions}\label{sec-conclusions}

...

>>>>

\section{Conclusions}\label{sec-conclusions}

We have investigated multiple build systems, showing how their properties are consequences of two implementation choices: what order you build in and whether you decide to rebuild. By first decomposing the pieces, we show how to recompose the pieces to find new points in the design space. In particular, a simple recombination leads to a design for a monadic cloud build system. Armed with that blueprint we hope to actually implement such a system as future work.
